### 2022-08-23 article.tex
Citations are surrounded by blocky brackets only

Genetic epidemiology is a field within biology that, among 
others, measures the spread of heritable traits,
as well as how these come to be.
For example, we know that lactose intolerance is, among others,
caused by producing too few lactose-degrading enzymes,
and is most commonly found in south-east asia.
The trait is caused by the genetic make-up, 
or genotype (see Table \ref{tab:definitions}), of a person.
The trait, also called phenotype, in this example is lactose intolerance,
but any human property, such as weight or height can be studied.
Genetic epidemiology studies the spread of genotypes and phenotypes,
as well as the relation between those two.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create gene panels, where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for a phenotype.

@@

The example study followed is a pseudorandomly selected paper
from \cite{ahsan2017relative}. The primary data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 (\cite{igl2010northern}). 
The approximately 1000 participants were initially mostly surveyed
about lifestyle (\cite{igl2010northern}) and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genotypes (\cite{johansson2013identification}),
(2) the phenotypes, which are the concentrations of certain proteins
in the blood (\cite{enroth2014strong,enroth2015effect}).

@@

A scientific claim must be reproducible before it is accepted
by the scientific community.
Reproducibility (i.e. to reproduce the same results) 
and replicability (i.e. to reconclude a conclusion)
are fundamental characteristics of scientific studies (\cite{patil2019visual}).
For computational science, it is relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In computational science, however, it appears that 
culture to reproduce results has been lost 
and to counter this, it has been suggested to make
reproducible research a minimal requirement for 
publication (\cite{peng2011reproducible}).

@@

The code of a computational experiment must be published
for an experiment to reproducible.
It is however (too) common that the original data
and/or code are unavailable (\cite{peng2021reproducible}).
One saddening example is described in \cite{haibe2020importance}, where
an algorithm that detects breast cancer better than a human expert,
was irreproducible.
In the case of genetic epidemiology, it is a given that the original
data cannot be published as this data is sensitive.
The code of a computational experiment, however, \emph{can} be
distributed without such problems.
It is, however, not code that is published, but an English
description of what that code does instead.
From a knowledge management's perspective,
to conserve an experiment,
it is the code that must be archived and not the
English description of what it does:
it is the code does the actual work.
Hence, from a knowledge management perspective, emphasis should be
put more on the preservation of code, as it is the most important actor
in an experiment.

@@

It is the paper that accompanies the code,
as it is the code that generates the results.
When humans are fallible and code gets bigger, the likelihood of
a mismatch between the English paper and the code increases.
We know that the most common errors are 
simple (and may legitimately fear that those simple errors are common) 
(\cite{baggerly2009deriving}),
and this example is not too far off from the
examples given in that paper.

@@

As it is the code that is the most important actor in an the experiment,
it should be treated as equally or more important than the scientific
paper describing what is does.
Regardless of this, most academic journals, 
when concerning computational experiments
only review the (usually) English description of the code
and make the review of code optional.
One notable exception is rOpenSci (\cite{ram2018community}),
which is an organisation that does peer-review academic code 
and facilitates the optional publication of a paper.
Although both code and paper are peer-reviewed,
most effort is spent on reviewing the code,
as can be seen in, for example, 
the \verb|mcbette| R package (\cite{bilderbeek2020mcbette}).

@@

The simplest metric to assess the quality of code is 
the amount of (single) lines of code (SLOC). 
Code that has a lower SLOC count, i.e. it is more concise,
is usually regarded as having a lower chance of containing bugs.
Additionally, following a style guide (e.g. \cite{style_guide}), 
improves software quality (\cite{fang2001}).

A more interesting metric is the cyclomatic complexity of the code.
The cyclomatic complexity is defined as the number of ways that
the code can be executed. 
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false.
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
(\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting})

A feature that deserves more attention
is that 'code must act as a teacher for future developers' (\cite{sadowski2018modern}).
Error handling is one of the mechanisms to do so.
A simple example is when one (e.g. a BSc student) 
needs to calculate the variance of a distribution.
Commonly one needs to call a function to do so, 
e.g. \verb|var| in R, that needs the values of the distribution
as input.
However, to calculate the variance of a distribution, at least two values
are required. 
When the variance of (zero or) only one value is requested,
an error message can helpfully indicate that the function 
needs at least two values.

@@

Code is sometimes published alongside a paper, 
in the form of a hosted website.
This way of publishing code gives additional options,
both for the authors of that code, as well as its users.
This practice has increased,
with the abstracts from articles in the academic journal
'Bioinformatics' referring to such a hosted website
has increased tenfold, going from below 5 percent in 2009 
to around 50 percent in 2017 (\cite{russell2018large}).

@@

There are multiple websites for hosting code, with BitBucket, GitHub,
GitLab and SourceForge being the most well-known.
GitHub is currenly the most popular website to host code in general,
with millions of users.
GitHub users can create dedicated websites (called 'repositories')
to upload code. Code hosts, such as GitHub, 
store that uploaded code and provide ways 
for visiters to download the code and/or interact with it.
The use of such code hosting websites
accommodates collaboration (\cite{perez2016ten}).
and improves transparency (\cite{gorgolewski2016practical}).
An example from the author is the website \cite{bbbqarticleissue157},
that hosts part of the code for the paper \cite{bilderbeek2022transmembrane},
as shown in Figure \ref{fig:bbbqarticleissue157}.

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{bbbq_article_issue_157.png}
  \caption{
    A typical GitHub repository (\cite{bbbqarticleissue157}), 
    hosting the code for a
    computational experiment, 
    as published with \cite{bilderbeek2022transmembrane}.
    Note that this repository has 72 commits, where a commit is a change
    to the code.
  }
  \label{fig:bbbqarticleissue157}
\end{figure}
\fi

@@

Hosted code commonly has a commit history. For example, GitHub
uses a version control system called \verb|git| (hence its name).
In practice, this means that when a change is made to the code,
a new version is created. In the case that the change was harmful,
one can go back to an earlier version and continue again from there.
It is a general recommendation to put version control
on all human-produced data (\cite{wilson2014best}).

@@

Hosted code improves transparency (\cite{gorgolewski2016practical}).
The history of changes, called the commit history, 
helps to assess the honesty in the reporting of the code. 
Ideally, the full version history of the code 
has been generated, by hosting the code directly at its inception.
However, in some repositories, only the final version of the code has been
posted, which means all historical information is lost.
From a code history, it can be observed, for example, 
how much statistical tests have really/actually been done 
and see if all are reported:
failing to reporting a (usually non-significant) statistical tests is 
one of the many degrees of freedom 
in p-value hacking (\cite{wicherts2016degrees}).

@@

Some code hosts, among others, GitHub, 
allows the researcher to start software scripts when uploading a new version.
To determine the reproducibility of the code, one such script should
run the code (on trivial data) to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediatly see which small change was the cause of this.
A formalized version of this workflow is called continuous integration (CI).
Using a CI service (such as GitHub Actions), is known to significantly 
increase the number of bugs exposed (\cite{vasilescu2015}) and increases
the speed at which new features are added (\cite{vasilescu2015}).
Using CI opens up the possibility to informally annotate features of
a repository. For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a LaTeX document (i.e. this paper)
could be built.

@@

Testing, in general, is an important mechanism to ensure
the correctness of code (an interesting example is \cite{rahman2020exploratory}
showing bugs in scientific software on the COVID-19 pandemic. 
It recommends more testing).
The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality (\cite{horgan1994,del1995correlation}). 
The non-profit organisation rOpenSci (\cite{ram2018community}),
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
Also here, CI opens up the possibility to informally annotate the
code coverage. For example, Figure \ref{fig:badge_codecov} shows a 
build badge that indicates that certain R package's code 
has a 100\% code coverage.
Hence, code coverage is important metadata, that can automatically
be acquired when using code hosting website.

@@

Most code hosts, among others, GitHub and GitLab,
automatically keep track of certain metrics.
One such metric is the amount of code each contributor has made.
In that way, authors involved in the writing of the software
are honestly acknowledged. Figure \ref{fig:daisie_contributors}
shows the number of commits the authors of DAISIE (\cite{etienne2020daisie})
have made to its code.
The distribution of contributed code per author 
is useful metadata, that can automatically
be acquired when using code hosting website.

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{daisie_contributors.png}
  \caption{
    An overview of the commits that multiple contributors have
    made, in this case, to the code of DAISIE (\cite{etienne2020daisie})
  }
  \label{fig:daisie_contributors}
\end{figure}
\fi

@@

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{issue_ldpred2.png}
  \caption{
    An overview of the issues for the code for
    LDpred2 (\cite{prive2020ldpred2}). Of the 310 issues, 300 have been closed.
    The first four issues have been replied to, as indicated by the
    talk balloon. The fifth issues has a label with the text 'feature request'
    to signal the issue type.
  }
  \label{fig:issue_ldpred2}
\end{figure}
\fi

@@

Albeit hosted code is vastly superior over raw code
in case of reproduciblity and correctness,
there are some problems with hosted code as well.
The most important one is that hosted code needs maintanance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
The degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here I will use the more modern (and in this context superior)
definition of software collapse (\cite{hinsen2019dealing}),
in which software fails due to dependencies on other 
software.
Where this paper recommends using older software foundations 
with multiple implementations (\cite{hinsen2019dealing}),
there is a modern solution that potentially has more longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data is
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy (\cite{zhang2016review,azencott2018machine}).

>>>>

Genetic epidemiology is a field within biology that, among 
others, measures the spread of heritable traits,
as well as how these come to be.
For example, we know that lactose intolerance is, among others,
caused by producing too few lactose-degrading enzymes,
and is most commonly found in south-east asia.
The trait is caused by the genetic make-up, 
or genotype 
(see Table \ref{tab:definitions} for the definitions of terms used in this paper), 
of a person.
The trait, also called phenotype (again, see Table \ref{tab:definitions}), 
in this example is lactose intolerance,
yet any human property, such as weight or height can be studied.
Genetic epidemiology studies the spread of genotypes and phenotypes,
as well as the relation between those two.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create gene panels, where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for the associated phenotype.

@@

The example study followed is a pseudorandomly selected paper
from \cite{ahsan2017relative}. The primary data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 \cite{igl2010northern}. 
The approximately 1000 participants were initially mostly surveyed
about lifestyle \cite{igl2010northern} and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genotypes \cite{johansson2013identification},
(2) the phenotypes, which are the concentrations of certain proteins
in the blood \cite{enroth2014strong,enroth2015effect}.

@@

A scientific claim must be reproducible before it is accepted
by the scientific community.
Reproducibility (i.e. to reproduce the same results) 
and replicability (i.e. to reconclude a conclusion)
are fundamental characteristics of scientific studies \cite{patil2019visual}.
For computational science, it is relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In computational science, however, it appears that 
culture to reproduce results has been lost 
and to counter this, it has been suggested to make
reproducible research a minimal requirement for 
publication \cite{peng2011reproducible}.

@@

The code of a computational experiment must be published
for an experiment to reproducible.
It is however (too) common that the original data
and/or code are unavailable \cite{peng2021reproducible}.
One saddening example is described in \cite{haibe2020importance}, where
an algorithm that detects breast cancer better than a human expert,
was irreproducible.
In the case of genetic epidemiology, it is a given that the original
data cannot be published as this data is sensitive.
The code of a computational experiment, however, \emph{can} be
distributed without such problems.
It is, however, not code that is published, but an English
description of what that code does instead.
From a knowledge management's perspective,
to conserve an experiment,
it is the code that must be archived and not the
English description of what it does:
it is the code does the actual work.
Hence, from a knowledge management perspective, emphasis should be
put more on the preservation of code, as it is the most important actor
in an experiment.

@@

It is the paper that accompanies the code,
as it is the code that generates the results.
When humans are fallible and code gets bigger, the likelihood of
a mismatch between the English paper and the code increases.
We know that the most common errors are 
simple (and may legitimately fear that those simple errors are common) 
\cite{baggerly2009deriving},
and this example is not too far off from the
examples given in that paper.

@@

As it is the code that is the most important actor in an the experiment,
it should be treated as equally or more important than the scientific
paper describing what is does.
Regardless of this, most academic journals, 
when concerning computational experiments
only review the (usually) English description of the code
and make the review of code optional.
One notable exception is rOpenSci \cite{ram2018community},
which is an organisation that does peer-review academic code 
and facilitates the optional publication of a paper.
Although both code and paper are peer-reviewed,
most effort is spent on reviewing the code,
as can be seen in, for example, 
the \verb|mcbette| R package \cite{bilderbeek2020mcbette}.

@@

The simplest metric to assess the quality of code is 
the amount of (single) lines of code (SLOC). 
Code that has a lower SLOC count, i.e. it is more concise,
is usually regarded as having a lower chance of containing bugs.
Additionally, following a style guide (e.g. \cite{style_guide}), 
improves software quality \cite{fang2001}.

A more interesting metric is the cyclomatic complexity of the code.
The cyclomatic complexity is defined as the number of ways that
the code can be executed. 
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false.
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting}

A feature that deserves more attention
is that 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
A simple example is when one (e.g. a BSc student) 
needs to calculate the variance of a distribution.
Commonly one needs to call a function to do so, 
e.g. \verb|var| in R, that needs the values of the distribution
as input.
However, to calculate the variance of a distribution, at least two values
are required. 
When the variance of (zero or) only one value is requested,
an error message can helpfully indicate that the function 
needs at least two values.

@@

Code is sometimes published alongside a paper, 
in the form of a hosted website.
This way of publishing code gives additional options,
both for the authors of that code, as well as its users.
This practice has increased,
with the abstracts from articles in the academic journal
'Bioinformatics' referring to such a hosted website
has increased tenfold, going from below 5 percent in 2009 
to around 50 percent in 2017 \cite{russell2018large}.

@@

There are multiple websites for hosting code, with BitBucket, GitHub,
GitLab and SourceForge being the most well-known.
GitHub is currenly the most popular website to host code in general,
with millions of users.
GitHub users can create dedicated websites (called 'repositories')
to upload code. Code hosts, such as GitHub, 
store that uploaded code and provide ways 
for visiters to download the code and/or interact with it.
The use of such code hosting websites
accommodates collaboration \cite{perez2016ten}.
and improves transparency \cite{gorgolewski2016practical}.
An example from the author is the website \cite{bbbqarticleissue157},
that hosts part of the code for the paper \cite{bilderbeek2022transmembrane},
as shown in Figure \ref{fig:bbbqarticleissue157}.

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{bbbq_article_issue_157.png}
  \caption{
    A typical GitHub repository \cite{bbbqarticleissue157}, 
    hosting the code for a
    computational experiment, 
    as published with \cite{bilderbeek2022transmembrane}.
    Note that this repository has 72 commits, where a commit is a change
    to the code.
  }
  \label{fig:bbbqarticleissue157}
\end{figure}
\fi

@@

Hosted code commonly has a commit history. For example, GitHub
uses a version control system called \verb|git| (hence its name).
In practice, this means that when a change is made to the code,
a new version is created. In the case that the change was harmful,
one can go back to an earlier version and continue again from there.
It is a general recommendation to put version control
on all human-produced data \cite{wilson2014best}.

@@

Hosted code improves transparency \cite{gorgolewski2016practical}.
The history of changes, called the commit history, 
helps to assess the honesty in the reporting of the code. 
Ideally, the full version history of the code 
has been generated, by hosting the code directly at its inception.
However, in some repositories, only the final version of the code has been
posted, which means all historical information is lost.
From a code history, it can be observed, for example, 
how much statistical tests have really/actually been done 
and see if all are reported:
failing to reporting a (usually non-significant) statistical tests is 
one of the many degrees of freedom 
in p-value hacking \cite{wicherts2016degrees}.

@@

Some code hosts, among others, GitHub, 
allows the researcher to start software scripts when uploading a new version.
To determine the reproducibility of the code, one such script should
run the code (on trivial data) to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediatly see which small change was the cause of this.
A formalized version of this workflow is called continuous integration (CI).
Using a CI service (such as GitHub Actions), is known to significantly 
increase the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
Using CI opens up the possibility to informally annotate features of
a repository. For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a LaTeX document (i.e. this paper)
could be built.

@@

Testing, in general, is an important mechanism to ensure
the correctness of code (an interesting example is \cite{rahman2020exploratory}
showing bugs in scientific software on the COVID-19 pandemic. 
It recommends more testing).
The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organisation rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
Also here, CI opens up the possibility to informally annotate the
code coverage. For example, Figure \ref{fig:badge_codecov} shows a 
build badge that indicates that certain R package's code 
has a 100\% code coverage.
Hence, code coverage is important metadata, that can automatically
be acquired when using code hosting website.

@@

Most code hosts, among others, GitHub and GitLab,
automatically keep track of certain metrics.
One such metric is the amount of code each contributor has made.
In that way, authors involved in the writing of the software
are honestly acknowledged. Figure \ref{fig:daisie_contributors}
shows the number of commits the authors of DAISIE \cite{etienne2020daisie}
have made to its code.
The distribution of contributed code per author 
is useful metadata, that can automatically
be acquired when using code hosting website.

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{daisie_contributors.png}
  \caption{
    An overview of the commits that multiple contributors have
    made, in this case, to the code of DAISIE \cite{etienne2020daisie}
  }
  \label{fig:daisie_contributors}
\end{figure}
\fi

@@

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{issue_ldpred2.png}
  \caption{
    An overview of the issues for the code for
    LDpred2 \cite{prive2020ldpred2}. Of the 310 issues, 300 have been closed.
    The first four issues have been replied to, as indicated by the
    talk balloon. The fifth issues has a label with the text 'feature request'
    to signal the issue type.
  }
  \label{fig:issue_ldpred2}
\end{figure}
\fi

@@

Albeit hosted code is vastly superior over raw code
in case of reproduciblity and correctness,
there are some problems with hosted code as well.
The most important one is that hosted code needs maintanance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
The degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here I will use the more modern (and in this context superior)
definition of software collapse \cite{hinsen2019dealing},
in which software fails due to dependencies on other 
software.
Where this paper recommends using older software foundations 
with multiple implementations \cite{hinsen2019dealing},
there is a modern solution that potentially has more longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data is
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy \cite{zhang2016review,azencott2018machine}.
